Higher Moments Distances from Robust Subspace for Botnet Detection

### 2019-05-02 docs/papers/2019JNCA_m-rpca/2019JNCA_m-rpca.tex
literature review OK

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. The botnet detection algorithms can also be improved by feature generation from aggregated traffic, considering that new features can make the data more discriminative for imbalanced learning problems. We also believe that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for botnet detection. Therefore, we propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggretation. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection for new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervisioned fashion and without the computational cost of new robust subspace learning for new observations.

We adopt a sampling method based on data aggregation for discriminative feature generation, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The autors recommend that none of the botnet families used in the training and crossvalidation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training. In contrast, we propose to evaluate each scenario of CTU-13 individually, in a semi-supervised approach that does not rely on training data with anomalies.

@@

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adoted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of location and scatter matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score distance and the orthogonal distance, calculate the thresholds and classify new observations accordingly.

In contrast to our proposal, ROBPCA and ROBPCA-AO rely on a previously defined cutoff value for classify anomalous observations, and does not require training.

Our robust PCA detector does not require having a perfect ground-truth for training; it just trains from the (eventually contaminated) background traffic. It is, in fact, an unsupervised learning algorithm. Moreover, the addition of feature selection allows matching the specificities of the traffic the detector is operating on, which can vary widely both in terms of the most frequent attacks and the relative weight of licit applications. Feature selection does require training with a labeled data set bu. One important strength of our anomaly detector is its adaptability to different network environments and traffic conditions

aims to reduce the sensitivity of PCA to outliers, by means of . In particular, RPCA allows for the careful teasing apart of sparse outliers so that the remaining low-dimensional approximation is faithful to the noise-free low-dimensional subspace describing the bulk of the raw data [6, 7, 20]. Specifically, RPCA splits a data matrix X into a low-rank matrix L and a sparse matrix S such that
X = L + S.

@@

Mahalanobis Distance is defined as		
	\begin{equation}\label{eq:eq01}
		d(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\hat{S}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\hat{S}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}.
	\end{equation}
$\boldsymbol{x}$ is a vector of a new observation, $\bar{\boldsymbol{x}}$ is the mean vector, also referred as location, of known observations and $\boldsymbol{\hat{S}}$ is the covariance matrix, also referred as scatter, of known observations.

Classical estimates can be so strongly affected by contamination that diagnostic tools, such as the Mahalanobis Distances, become unable to detect the outliers. Therefore, we need reliable and robust estimators that can resist outliers contaminated data. Minimum Covariance Determinant (MCD) \cite{rousseeuw1984mcd} is a robust estimator commonly used for computing Mahalanobis Distance in order to detect anomalies.

\begin{figure}[h!]
     \centering
     \includegraphics[width=11cm]{figures/mahalanobis_robust.png}
     \caption{Mahalanobis Distance for non-robust and robust estimates.}
     \label{fig:fig04}
\end{figure}

\begin{figure}[h!]
     \centering
     \includegraphics[width=8cm]{figures/mahalanobis_distances01.png}
     \caption{Mahalanobis Distance for non-robust and robust estimates.}
     \label{fig:fig05}
\end{figure}

The MCD estimator is one of the first affine equivariant and highly robust estimators of multivariate location and scatter.

Affine equivariant means that when the data are translated or subjected to a linear transformation, the resulting location and scatter will transform accordingly \cite{rousseeuw1984mcd,rousseeuw1999fastmcd}.

Its objective is to find $h$ observations (out of $n$) whose covariance matrix has the lowest determinant.

The lowest determinant means the lowest distance generalized variance and high distance between the largest eigenvalue and the remaining [ref?].

MCD depends on random sampling, therefore it is not deterministic.

MCD has been applied in numerous fields (medicine, finance, image analysis, and chemistry) and been used to develop many robust multivariate techniques (such as Robust Principal Component Analysis (RPCA), factor analysis, and multiple regression).

The classical MCD has rarely been applied because it is hard to compute, since it requires the evaluation of all $\binom{n}{h}$ subsets of size $h$. % binomial of n over p

Its main use has started since the construction of the computationally efficient Fast-MCD algorithm \cite{rousseeuw1999fastmcd}.

Fast-MCD is an iterative, approximate and resampling algorithm for the MCD.

It starts by random initial subset of size $p+1$ and performs concentration steps (C-steps) yielding consecutive h-subsets with decreasing covariance matrix determinant.

Only two C-steps are applied to each initial subset, and the 10 results with lowest determinant are kept.

C-steps are carried out until convergence and the best solution is kept.

Fast-MCD also is affine equivariant and not deterministic.

\textbf{Theorem 1: C-Step.}

Consider a data set $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ where $\boldsymbol{X} = \{\boldsymbol{x}_1,...,\boldsymbol{x}_n\}$ of $p$-variate observations. Let $\boldsymbol{h} \subset \{1,...,n\}$ with $|\boldsymbol{h}| = h$, where $\boldsymbol{h}$ is a subset of $h$ indexes from $n$ observations.

Let the mean $\bar{\boldsymbol{x}} \in \mathbb{R}^{1 \times p}$ be

\begin{equation}\label{eq:eq02}
	\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{j\in \boldsymbol{h}} \boldsymbol{x}_j, 
\end{equation}

the covariance matrix $\boldsymbol{\hat{S}} \in \mathbb{R}^{p \times p}$ be

\begin{equation}\label{eq:eq03}
	\boldsymbol{\hat{S}} = \displaystyle\frac{1}{h}\displaystyle\sum_{j\in \boldsymbol{h}} (\boldsymbol{x}_j - \bar{\boldsymbol{x}})(\boldsymbol{x}_j - \bar{\boldsymbol{x}})^\prime,
\end{equation}

and the relative distance be

\begin{equation}\label{eq:eq04}
	\boldsymbol{d}(i) = \sqrt{(\boldsymbol{x}_i - \bar{\boldsymbol{x}}) \boldsymbol{\hat{S}}^{-1}(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime}, 
\end{equation}

for $i = 1,...,n$.

Let $\boldsymbol{h}_1, \bar{\boldsymbol{x}}_1$ and $\boldsymbol{\hat{S}}_1$ and be the selected observations, the location and scatter of the initial subset, and $\boldsymbol{d}_1(i)$ be the initial distances for $i = 1,...,n$.

Now take $\boldsymbol{h}_2$ containing the indexes of lowest distances, such that $\{\boldsymbol{d}_1(i); i \in \boldsymbol{h}_2\} = \{(\boldsymbol{d}_1)_{1:n},...,(\boldsymbol{d}_1)_{h:n}\}$, where $(\boldsymbol{d}_1)_{1:n} \leq (\boldsymbol{d}_1)_{2:n} \leq ... \leq (\boldsymbol{d}_1)_{n:n}$ are the ordered distances, and compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$.

Then, $det(\boldsymbol{\hat{S}}_2) \leq det(\boldsymbol{\hat{S}}_1)$ with equality if and only if $\boldsymbol{\bar{x}}_2 = \boldsymbol{\bar{x}}_1$ and $\boldsymbol{\bar{S}}_2 = \boldsymbol{\hat{S}}_1$

The proof is given by Rousseeuw and Driessen \cite{rousseeuw1999fastmcd}.

\textbf{The C-step can be algorithmically described as follows:}
\begin{algorithm}
	\label{alg:alg01}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
	Given the initial $h-$subset $\boldsymbol{h}_1$ of indexes or the pair $(\boldsymbol{\bar{x}}_1, \boldsymbol{\hat{S}}_1)$\;
	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
		\If{$\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ exist}{
			Put $\boldsymbol{\bar{x}}_1 = \boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_1 = \boldsymbol{\hat{S}}_2$\;
		}
		Compute the distances $\boldsymbol{d}_1( i )$ for $i = 1,..., n$\;
		Sort $\boldsymbol{d}_1( i )$, which yields an index $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_1(\pi(n))$\;
		Put $\boldsymbol{h}_2 = \{ \pi(1), \pi(2), ..., \pi(h)\}$\;
		Compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$\;
		Compute $det(\boldsymbol{\hat{S}}_2)$ and $det(\boldsymbol{\hat{S}}_1)$\;			
	}
	\caption{C-Step}
\end{algorithm}

\textbf{Theorem 1} thus provides a partial idea for an algorithm:

\textit{Take many initial choices of $\boldsymbol{h}_1$ and apply C-steps to each until convergence (don't have determinant decreasing), and keep the result with lowest determinant}.

\textbf{Initial subset can be algorithmically described as:}
\begin{algorithm}
	\label{alg:alg02}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{h}_1$}
	Given a data set $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ with $p$-variate observations\;
	Draw a random $(p + 1)$-subset $\boldsymbol{L}$\;
	Compute $\boldsymbol{\bar{x}}_1$ and $\boldsymbol{\hat{S}}_1$ of $\boldsymbol{L}$ and $det(\boldsymbol{\hat{S}}_1)$\;
	\If{$det(\boldsymbol{\hat{S}}_1) \leq 0$}{
		Add another random observation into $\boldsymbol{L}$\;
		Compute $\boldsymbol{\bar{x}}_1$ and $\boldsymbol{\hat{S}}_1$ of $\boldsymbol{L}$ and $det(\boldsymbol{\hat{S}}_1)$\;
	}
	Compute $\boldsymbol{d}_1^2(i) = (\boldsymbol{x}_i - \boldsymbol{\bar{x}}_1)^\prime \boldsymbol{\hat{S}}_1^{-1}(\boldsymbol{X}_i - \boldsymbol{\bar{x}}_0)$ for $i = 1, ..., n$\;
	Sort them into $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_0(\pi(n))$\;
	Put $\boldsymbol{h}_1 = \{\pi(1), ..., \pi(h)\}$\;
	\caption{Constructing the initial subset}
\end{algorithm}

\begin{algorithm}
	\label{alg:alg03}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
	Given $\boldsymbol{X} \in \mathbb{R}^{n \times p}$, $h < n$, $p \geq 2$ and $n \leq 600$\;
	\For{500 times}{
		Construct initial $h$-subset $\boldsymbol{h}_1$ (Algorithm \ref{alg:alg02})\;
		Carry out two C-steps (Algorithm \ref{alg:alg01}) resulting $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$\;
	}		
	Put $\boldsymbol{h}_1$ as the 10 lowest $det(\boldsymbol{\hat{S}_2})$ computed above\;
	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
		Carry out C-steps (Algorithm \ref{alg:alg01})\;
	}
	\caption{Fast-MCD when $n \leq 600$}
\end{algorithm}

\begin{algorithm}
	\label{alg:alg04}
	\scriptsize
	\SetAlgoLined
	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
	Given $\boldsymbol{X} \in \mathbb{R}^{n \times p}$, $h < n$, $p \geq 2$ and $n > 600$\;
	Construct up to 5 disjoint random subsets of size $n_{sub} = 300$\;
	\For{each subset, repeat 100 times}{
		Construct initial $\boldsymbol{h}_1$ of size $h_{sub} = [n_{sub}(h/n)]$\;
		Carry out two C-steps of $n_{sub}$ and $h_{sub}$\;
		Keep the 10 best $(\boldsymbol{\bar{x}}_{sub}, \boldsymbol{\hat{S}}_{sub})$\;
	}		
	Pool the subsets, yielding the merged set of $n_{merged} = 1500$\;
	\For{50 best $(\boldsymbol{\bar{x}}_{sub}, \boldsymbol{\hat{S}}_{sub})$}{
		Carry out two C-steps of $n_{merged}$ and $h_{merged} = [n_{merged}(h/n)]$\;
		Keep the 10 best $(\boldsymbol{\bar{x}}_{merged}, \boldsymbol{\hat{S}}_{merged})$\;
	}
	\For{the full data set and $m_{full}$ best results}{
		Take several C-steps, using $n$ and $h$\;
		Keep the best final result $(\boldsymbol{\bar{x}}_{full}, \boldsymbol{\hat{S}}_{full})$\;
	}
	\caption{Fast-MCD when $n > 600$}
\end{algorithm}

% feature generation and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Using entropy as a summarization tool, we show that the analysis of feature distributions leads to significant advances on two fronts: (1) it enables highly sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, and (2) it enables automatic classification of anomalies via unsupervised learning. Anomaly diagnosis framework, comprising both an extension of the subspace method [23] to accommodate multiple data types, and an unsupervised classification technique using simple clustering algorithms. \textbf{To explain feature selection and aggregation}. Goldberg and Shan \cite{goldberg2015importance} argue that processing monitored signals into features is a more fruitful way to build anomaly detection systems than devising ever more complex statistical tests, based on experience and  on putblished real-life examples of anomaly detectors in use at eBay.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highligh that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis. Therefore, learning methods for imbalanced and skewed data have been atracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}. Skewness and the kurtosis are very sensitive outlier detectors, considering that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

Our proposal is ...

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively. We define the operator $\rm diag(\cdot)$ that returns the vector of the main diagonal of a given matrix, the operator $\rightarrow$, which denotes the deletion of a given element from a set and the operator $\#$, that returns the rank of a matrix, and the operator $\sim$ that sorts the elements of a vector in ascending order.

This section also presents a description of the data model and data aggregation of the CTU-13 data set, in Subsection \ref{sec:CTU-13} discusses the use of the CTU-13 data set for evaluation of the proposed approach.

@@

@@

@@

	Given a contaminated data set, split it into normal (60\%), cross validation (20\% normal + 50\% anomalous) and test (20\% normal + 50\% anomalous).
	Select the range of contamination that shall be evaluated and perform cross validation for each contamination. 
	For each contamination $c$:

		Estimate $\boldsymbol{\bar{x}}_{normal}$ and $\boldsymbol{\hat{S}}_{normal}$ from normal data.
		Predict anomalies for cross validation data using the $c$, $\boldsymbol{\bar{x}}_{normal}$ and $\boldsymbol{\hat{S}}_{normal}$.
		Select the best F-measure by cross validation and get the best $c$.

@@

>>>>

We believe that the skewness of anomalous and normal traffic can highlight features for improving anomaly detection in imbalanced data. The botnet detection algorithms can also be improved by feature engineering from aggregated traffic, considering that new features can make the data more discriminative for imbalanced learning problems. We also believe that the distance between robust estimates of normal traffic and anomalies can highlight discrepancies and be used for botnet detection. Therefore, we propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggretation. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection for new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervisioned fashion and without the computational cost of new robust subspace learning for new observations.

We adopt a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

The objective of Garcia \emph{et al.} \cite{garcia2014empirical} is to compare three botnet detection methods using a simple and reproducible methodology, a good data set and a new error metric. This paper evaluates some data sets for network anomaly detection, surveys some approaches for botnet detection, proposes two methods (BClus and CAMNEP) for botnet detection and compare results to BotHunter \cite{gu2007bothunter}. Considering the lack of available labeled data sets, Garcia \emph{et al.} \cite{garcia2014empirical} proposes the CTU-13 data set, which is composed by normal and background labeled data, in an imbalanced distribution like in a real network. The autors recommend that none of the botnet families used in the training and crossvalidation data set should be used in the testing data set, aiming to ensure that the evaluated methods can generalize and detect new behaviors. Adopting the training and testing approach of \cite{garcia2014empirical}, some botnet malwares wouldn't be tested, since in the author's proposal some botnets are present only for training.

@@

RPCA is been mainly applied to computer vision, in problems of robust subspace tracking and robust subspace recovery. However, RPCA has also been adoted for general outlier detection \cite{hubert2005robpca,hubert2009robustskewed,cherapanamjeri2017thresholding,zhou2017anomaly,NetflixSurus} and for anomaly detection on network traffic \cite{pascoal2012robust}. ROBPCA \cite{hubert2005robpca} intends to identify outliers using PCA from robust estimates of location and scatter matrix, to reduce the data dimensions and plotting the orthogonal distances versus the robust score distances, to flag an outlier map. However, ROBPCA flags many points as outlying when the original data is skewed. Therefore, Hubert \emph{et al.} \cite{hubert2009robustskewed} proposed ROBPCA-AO, which improves ROBPCA for problems with skewed data, by means of an adjusted outlyingness based on robust skewness. Pascoal \emph{et al.} \cite{pascoal2012robust} proposed an approach based on a robust mutual information estimator for feature selection and based on RPCA for outlier detection in internet traffic. The anomaly detection proposed by Pascoal \emph{et al.} is an unsupervised approach that estimate the first $k$ robust principal components, calculate the score and the orthogonal distances, calculate the thresholds and classify new observations accordingly.

@@

% feature engineering and aggregation
Lakhina \emph{et al.} \cite{lakhina2005mining} argue that the distributions of packet features (IP addresses and ports) observed in flow traces reveals both the presence and the structure of a wide range of anomalies. Lakhina \emph{et al.} show that the analysis of feature distributions leads to significant advances on sensitive detection of a wide range of anomalies, augmenting detections by volume-based methods, and leads gains on automatic classification of anomalies via unsupervised learning. Goldberg and Shan \cite{goldberg2015importance} argue that processing monitored signals into features is a more fruitful way to build anomaly detection systems than devising ever more complex statistical tests, based on experience and on putblished real-life examples of anomaly detectors in use at eBay.

Benson \emph{et al.} \cite{benson2010network} conducted an empirical study of the network traffic in 10 data centers belonging to three different types of organizations, including university, enterprise, and cloud data centers. Findings of Benson \emph{et al.} indicate that certain positive skewed and heavy-tailed distributions can model data center switch traffic, and highlights a difference between the data center environment and the wide area network, where the long-tailed Pareto distribution typically shows the best fit \cite{benson2010network}. This findings highligh that the skeweness and heavy-tailed distributions can impact algorithms that rely on guassian distributed data, as well as it can reveal characteristics that can be exploited in order to obtain accurate classifiers for network traffic analysis. Considering that learning methods for imbalanced and skewed data have been atracted attention of researchers \cite{Phua2004minority,hubert2009robustskewed}, higher moments have been adopted for outlier detection and discriminative analysis. Skewness and the kurtosis are higher moments that can contribute to sensitive outlier detectors, taking into account that the properties of the kurtosis can be used to identify clusters or outliers \cite{pena2010eigenvectors}, while skewness has been used as adjust factor for outlier detection \cite{hubert2009robustskewed}.

We propose an approach based on higher moments distances from robust subspace for botnet detection, applied to new features generated by flow aggretation from the CTU-13 data set. The proposed approach relies on a robust subspace of normal traffic for estimating higher moments. The anomaly detection for new observations evaluate the Mahalanobis distance between the robust higher moments and the higher moments of new observations, in a semi-supervisioned fashion and without the computational cost of new robust subspace learning for new observations.

We adopt a sampling method based on data aggregation for discriminative feature engineering, considering that data aggregation is a widely adoted approach for feature engineering \cite{garcia2014empirical, chandrashekar2014survey,acarali2016survey} and network anomaly detection \cite{lakhina2005mining, callegari2011novel, vieira2017model}. Finally, We evaluate the accuracy of the proposed approach for botnet detecion on the CTU-13 data set \cite{garcia2014empirical}, which is a large data set of normal, background and botnet traffic that has been adopted to deal with the lack of up-to-date real-world data sets for anomaly detection systems \cite{osanaiye2016distributed}.

@@

In this paper, scalars are denoted by italic letters (\emph{a, b, A, B, $α$, $β$}), vectors by lowercase bold letters (\textbf{a, b}), matrices by uppercase bold letters (\textbf{A, B}), and $a_i,_j$ denotes the (\emph{i, j}) elements of the matrix \textbf{A}. The superscripts \textsuperscript{T} and \textsuperscript{-1} are used for matrix transposition and matrix inversion, respectively.

This section also presents a description of the data model in Subsection \ref{sec:CTU-13} and in Subsection \ref{sec:feat_eng} we describe the data aggregation and feature engineering adopted for the CTU-13.

@@

\subsection{feature engineering}
\label{sec:feat_eng}

@@

Experimental evaluation compares our proposal to widely adopted algorithms for anomaly detection based on clustering and statisctical approaches, which are K-means and Gaussian Mixture Model (GMM) \cite{gaddam2007kmeans,moustafa2019holistic}, respectively. We also evaluate the results of ROBPCA \cite{hubert2005robpca}, wich is a method that also relies on robust estimates with adjusted outlyingness based on robust skewness. In contrast to Garcia \emph{et al.} \cite{garcia2014empirical}, we propose to evaluate each scenario of CTU-13 individually, in a semi-supervised approach that does not rely on training data with anomalies.

@@

Given a contaminated data set, split it into normal (60\%), cross validation (20\% normal + 50\% anomalous) and test (20\% normal + 50\% anomalous).

Select the range of contamination that shall be evaluated and perform cross validation for each contamination.

For each contamination $c$:

Estimate $\boldsymbol{\bar{x}}_{normal}$ and $\boldsymbol{\hat{S}}_{normal}$ from normal data.

Predict anomalies for cross validation data using the $c$, $\boldsymbol{\bar{x}}_{normal}$ and $\boldsymbol{\hat{S}}_{normal}$.

Select the best F-measure by cross validation and get the best $c$.

@@

% Mahalanobis Distance is defined as		
% 	\begin{equation}\label{eq:eq01}
% 		d(\boldsymbol{x},\bar{\boldsymbol{x}}, \boldsymbol{\hat{S}}) = \sqrt{(\boldsymbol{x} - \bar{\boldsymbol{x}}) \boldsymbol{\hat{S}}^{-1}(\boldsymbol{x} - \bar{\boldsymbol{x}})^\prime}.
% 	\end{equation}
% $\boldsymbol{x}$ is a vector of a new observation, $\bar{\boldsymbol{x}}$ is the mean vector, also referred as location, of known observations and $\boldsymbol{\hat{S}}$ is the covariance matrix, also referred as scatter, of known observations.

% Classical estimates can be so strongly affected by contamination that diagnostic tools, such as the Mahalanobis Distances, become unable to detect the outliers. Therefore, we need reliable and robust estimators that can resist outliers contaminated data. Minimum Covariance Determinant (MCD) \cite{rousseeuw1984mcd} is a robust estimator commonly used for computing Mahalanobis Distance in order to detect anomalies.

% \begin{figure}[h!]
%      \centering
%      \includegraphics[width=11cm]{figures/mahalanobis_robust.png}
%      \caption{Mahalanobis Distance for non-robust and robust estimates.}
%      \label{fig:fig04}
% \end{figure}

% \begin{figure}[h!]
%      \centering
%      \includegraphics[width=8cm]{figures/mahalanobis_distances01.png}
%      \caption{Mahalanobis Distance for non-robust and robust estimates.}
%      \label{fig:fig05}
% \end{figure}

% The MCD estimator is one of the first affine equivariant and highly robust estimators of multivariate location and scatter.

% Affine equivariant means that when the data are translated or subjected to a linear transformation, the resulting location and scatter will transform accordingly \cite{rousseeuw1984mcd,rousseeuw1999fastmcd}.

% Its objective is to find $h$ observations (out of $n$) whose covariance matrix has the lowest determinant.

% The lowest determinant means the lowest distance generalized variance and high distance between the largest eigenvalue and the remaining [ref?].

% MCD depends on random sampling, therefore it is not deterministic.

% MCD has been applied in numerous fields (medicine, finance, image analysis, and chemistry) and been used to develop many robust multivariate techniques (such as Robust Principal Component Analysis (RPCA), factor analysis, and multiple regression).

% The classical MCD has rarely been applied because it is hard to compute, since it requires the evaluation of all $\binom{n}{h}$ subsets of size $h$. % binomial of n over p

% Its main use has started since the construction of the computationally efficient Fast-MCD algorithm \cite{rousseeuw1999fastmcd}.

% Fast-MCD is an iterative, approximate and resampling algorithm for the MCD.

% It starts by random initial subset of size $p+1$ and performs concentration steps (C-steps) yielding consecutive h-subsets with decreasing covariance matrix determinant.

% Only two C-steps are applied to each initial subset, and the 10 results with lowest determinant are kept.

% C-steps are carried out until convergence and the best solution is kept.

% Fast-MCD also is affine equivariant and not deterministic.

% \textbf{Theorem 1: C-Step.}

% Consider a data set $\boldsymbol{X} \in \mathbb{R}^{n \times p}$ where $\boldsymbol{X} = \{\boldsymbol{x}_1,...,\boldsymbol{x}_n\}$ of $p$-variate observations. Let $\boldsymbol{h} \subset \{1,...,n\}$ with $|\boldsymbol{h}| = h$, where $\boldsymbol{h}$ is a subset of $h$ indexes from $n$ observations.

% Let the mean $\bar{\boldsymbol{x}} \in \mathbb{R}^{1 \times p}$ be

% \begin{equation}\label{eq:eq02}
% 	\bar{\boldsymbol{x}} = \displaystyle\frac{1}{h}\displaystyle\sum_{j\in \boldsymbol{h}} \boldsymbol{x}_j, 
% \end{equation}

% the covariance matrix $\boldsymbol{\hat{S}} \in \mathbb{R}^{p \times p}$ be

% \begin{equation}\label{eq:eq03}
% 	\boldsymbol{\hat{S}} = \displaystyle\frac{1}{h}\displaystyle\sum_{j\in \boldsymbol{h}} (\boldsymbol{x}_j - \bar{\boldsymbol{x}})(\boldsymbol{x}_j - \bar{\boldsymbol{x}})^\prime,
% \end{equation}

% and the relative distance be

% \begin{equation}\label{eq:eq04}
% 	\boldsymbol{d}(i) = \sqrt{(\boldsymbol{x}_i - \bar{\boldsymbol{x}}) \boldsymbol{\hat{S}}^{-1}(\boldsymbol{x}_i - \bar{\boldsymbol{x}})^\prime}, 
% \end{equation}

% for $i = 1,...,n$.

% Let $\boldsymbol{h}_1, \bar{\boldsymbol{x}}_1$ and $\boldsymbol{\hat{S}}_1$ and be the selected observations, the location and scatter of the initial subset, and $\boldsymbol{d}_1(i)$ be the initial distances for $i = 1,...,n$.

% Now take $\boldsymbol{h}_2$ containing the indexes of lowest distances, such that $\{\boldsymbol{d}_1(i); i \in \boldsymbol{h}_2\} = \{(\boldsymbol{d}_1)_{1:n},...,(\boldsymbol{d}_1)_{h:n}\}$, where $(\boldsymbol{d}_1)_{1:n} \leq (\boldsymbol{d}_1)_{2:n} \leq ... \leq (\boldsymbol{d}_1)_{n:n}$ are the ordered distances, and compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$.

% Then, $det(\boldsymbol{\hat{S}}_2) \leq det(\boldsymbol{\hat{S}}_1)$ with equality if and only if $\boldsymbol{\bar{x}}_2 = \boldsymbol{\bar{x}}_1$ and $\boldsymbol{\bar{S}}_2 = \boldsymbol{\hat{S}}_1$

% The proof is given by Rousseeuw and Driessen \cite{rousseeuw1999fastmcd}.

% \textbf{The C-step can be algorithmically described as follows:}
% \begin{algorithm}
% 	\label{alg:alg01}
% 	\scriptsize
% 	\SetAlgoLined
% 	\KwResult{$\boldsymbol{\bar{x}}_2$, $\boldsymbol{\hat{S}}_2$}
% 	Given the initial $h-$subset $\boldsymbol{h}_1$ of indexes or the pair $(\boldsymbol{\bar{x}}_1, \boldsymbol{\hat{S}}_1)$\;
% 	\While{$det(\boldsymbol{\hat{S}}_2) < det(\boldsymbol{\hat{S}}_1)$}{
% 		\If{$\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ exist}{
% 			Put $\boldsymbol{\bar{x}}_1 = \boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_1 = \boldsymbol{\hat{S}}_2$\;
% 		}
% 		Compute the distances $\boldsymbol{d}_1( i )$ for $i = 1,..., n$\;
% 		Sort $\boldsymbol{d}_1( i )$, which yields an index $\boldsymbol{d}_1(\pi(1)) \leq ... \leq \boldsymbol{d}_1(\pi(n))$\;
% 		Put $\boldsymbol{h}_2 = \{ \pi(1), \pi(2), ..., \pi(h)\}$\;
% 		Compute $\boldsymbol{\bar{x}}_2$ and $\boldsymbol{\hat{S}}_2$ for $\boldsymbol{h}_2$\;
% 		Compute $det(\boldsymbol{\hat{S}}_2)$ and $det(\boldsymbol{\hat{S}}_1)$\;			
% 	}
% 	\caption{C-Step}
% \end{algorithm}

% \textbf{Theorem 1} thus provides a partial idea for an algorithm:

% \textit{Take many initial choices of $\boldsymbol{h}_1$ and apply C-steps to each until convergence (don't have determinant decreasing), and keep the result with lowest determinant}.
